---
title: "TCCIP data regionally extracting"
output: html_notebook
---
# 概述
將TCCIP的資料依據研究範圍挑選出來，再進行後續計算。
# 0. package
```{r,message=FALSE}
library(data.table)
library(sf)
library(tidyverse)
library(readr) # read the file in .zip
library(parallel)  
library(SPEI)
library(magick)
w_path <- "D:/climate_data/TCCIP/日資料/1960_2022/"
pro_path <-"G:/我的雲端硬碟/研究室計畫/2024/玉管處/2024/" 
```

# 1. extract the TCCIP point in the specific region.
Due to the weather data in TCCIP is presented in a tabular format. We need to filter the point which was located in the national park.

```{r}
region <- st_read("rawdata/layer/Yushan_national_park_region.shp")
tccip_p <- fread(paste0(w_path,"觀測_日資料_臺灣_平均溫/觀測_日資料_臺灣_平均溫_1960.csv"))
tccip_p <- tccip_p[,.(LON,LAT)]
tccip_p <- st_as_sf(tccip_p, coords = c("LON", "LAT"), crs = 4326)
class(tccip_p)
p_npr <- st_intersection(tccip_p,region)
p_coord <- st_coordinates(p_npr) %>% as.data.table()
setnames(p_coord,c("X","Y"),c("LON","LAT"))
p_coord[,c_id:=paste0(LON,LAT)]
```

# 2. extract the TCCIP weather data by point in NP
pass the p_coord to extract the weather data from TCCIP table
## 2.1 established the function
```{r}

tccip_region_ext <- function(i,var,wpath,coord){
  require(data.table)
  dt <- fread(paste0(wpath,"觀測_日資料_臺灣_",var,"/觀測_日資料_臺灣_",var,"_",i,".csv"))
  dt[,c_id:=paste0(LON,LAT)]
  dt_np <-dt[coord,on=.(c_id)]
  dt_np[,c_id:=NULL]
  dt_t <- melt(dt_np,id.vars=c("LON","LAT"),variable.name="date_t")
  dt_t[,date:=as.Date(date_t,format="%Y%m%d")][
    ,c("year","month"):=.(year(date),month(date))]
  dt_t <- dt_t[!is.na(date)]
  if (grepl("溫",var)){
    dt_m <- dt_t[,.(value=mean(value,na.rm=TRUE)),by=.(LON,LAT,year,month)]
  }else{
    dt_m <- dt_t[,.(value=sum(value,na.rm=TRUE)),by=.(LON,LAT,year,month)]
  }
  return(dt_m)
}

```
## 2.2 data processing
```{r}
year <- 1995:2022
core <- makeCluster(10)
temp_mean <-parLapply(core,year,tccip_region_ext,var="平均溫",wpath=w_path,coord=p_coord)
temp_max <-parLapply(core,year,tccip_region_ext,var="最高溫",wpath=w_path,coord=p_coord)
temp_min <-parLapply(core,year,tccip_region_ext,var="最低溫",wpath=w_path,coord=p_coord)
pre <-parLapply(core,year,tccip_region_ext,var="降雨量",wpath=w_path,coord=p_coord)
stopCluster(core)

clm_all <- list(temp_mean,temp_max,temp_min,pre)

```

## 2.3 calculated the weather index

```{r}

clm_all <- lapply(clm_all,rbindlist)
clm_all <- lapply(clm_all,function(clm_dt){
  clm_dt[,wkt:=paste("POLYGON ((",
                   LON-0.005,LAT-0.005,",",
                   LON+0.005,LAT-0.005,",",
                   LON+0.005,LAT+0.005,",",
                   LON-0.005,LAT+0.005,",",
                   LON-0.005,LAT-0.005,"))",
                   sep=" ")]
  return(clm_dt)
})

clm_Y <- lapply(clm_all,function(clm_dt){
  yr <- clm_dt[,.(value=mean(value)),by=.(LON,LAT,year,wkt)]
  yr[,v_mean:=mean(value),by=.(wkt)]
  yr[,v_sd:=sd(value),by=.(wkt)]
  yr[,st_value:=(value-v_mean)/v_sd]
  return(yr)
})
clm_Y[[4]][,value:=value*12]
y_var <- c("mean_T","max_T"
           ,"min_T","PRE")
lapply(1:3,function(i){
  dt <- clm_Y[[i]]
  var <- y_var[i]
  dir.create(paste0("result/plot/yearly_",var))
  lapply(1995:2022,function(y){
    dt_y <- dt[year==y]
    dt_y <-st_as_sf(dt_y,wkt ="wkt",crs=4326) 
    p <- ggplot()+
          geom_sf(data=dt_y,aes(fill=st_value))+
          labs(fill="TI",title =y)+
          scale_fill_gradient2(
            high ="#A62621",
            mid = "gray80",
            low = "#3C56A6",
            midpoint = 0,
            limits=c(-4,4)
          )
    ggsave(paste0("result/plot/yearly_",var,"/",var,"_",y,".jpeg"),
           height = 5,width = 8,dpi=600,plot=p)
  })
})

y_var <- c("mean_T","max_T"
           ,"min_T","PRE")
lapply(1:3,function(i){
  dt <- clm_Y[[i]]
  var <- y_var[i]
  dir.create(paste0("result/plot/yearly_org_",var))
  lapply(1995:2022,function(y){
    dt_y <- dt[year==y]
    dt_y <-st_as_sf(dt_y,wkt ="wkt",crs=4326) 
    p <- ggplot()+
          geom_sf(data=dt_y,aes(fill=value))+
          labs(fill=var,title =y)+
          scale_fill_gradient(
            high ="#A62621",
            low = "#3C56A6",
            limits=c(min(dt$value),max(dt$value))
          )
    ggsave(paste0("result/plot/yearly_org_",var,"/org",var,"_",y,".jpeg"),
           height = 5,width = 8,dpi=600,plot=p)
  })
})
PRE_Y <- clm_Y[[4]]
### deal with precipitation
  lapply(1995:2022,function(y){
    dt_y <- PRE_Y[year==y]
    dt_y <-st_as_sf(dt_y,wkt ="wkt",crs=4326) 
    p <- ggplot()+
          geom_sf(data=dt_y,aes(fill=st_value))+
          labs(fill="rPRE",title =y)+
          scale_fill_gradient2(
            high ="#40A928",
            mid = "gray80",
            low = "#DEB91B",
            midpoint = 0,
            limits=c(-4,4)
          )
    ggsave(paste0("result/plot/yearly_PRE/PRE_",y,".jpeg"),
           height = 5,width = 8,dpi=600,plot=p)
  })
```
 
## 2.4 save the data
 
```{r}
y_var <- c("mean_T","max_T"
           ,"min_T","PRE")
### save the total weather data table
lapply(1:4,function(i){
  dt <- clm_all[[i]]
  var <- y_var[i]
  fwrite(dt,paste0(pro_path,"rawdata/weather_table/basic_data_",var,".csv"))
  })
```
 
 