---
title: "TCCIP data regionally extracting"
output: html_notebook
---
# 概述
將TCCIP的資料依據研究範圍挑選出來，再進行後續計算。
# 0. package
```{r,message=FALSE}
library(data.table)
library(sf)
library(tidyverse)
library(readr) # read the file in .zip
library(parallel)  
library(SPEI)
library(magick)
library(patchwork)
w_path <- "D:/climate_data/TCCIP/日資料/1960_2022/"
pro_path <-"G:/我的雲端硬碟/研究室計畫/2024/玉管處/2025/" 

```

# 1. extract the TCCIP point in the specific region.
Due to the weather data in TCCIP is presented in a tabular format. We need to filter the point which was located in the national park.

```{r}

region <- st_read(paste0(pro_path,"rawdata/layer/玉山NP(4通)範圍界1130312/Yushan_NP_region.shp"))

tccip_p <- fread(paste0(w_path,"觀測_日資料_臺灣_平均溫/觀測_日資料_臺灣_平均溫_1960.csv"))
tccip_p <- tccip_p[,.(LON,LAT)]
tccip_p <- st_as_sf(tccip_p, coords = c("LON", "LAT"), crs = 4326)
class(tccip_p)
p_npr <- st_intersection(tccip_p,region)
p_coord <- st_coordinates(p_npr) %>% as.data.table()
setnames(p_coord,c("X","Y"),c("LON","LAT"))
p_coord[,c_id:=paste0(LON,LAT)]
```

# 2. extract the TCCIP weather data by point in NP
pass the p_coord to extract the weather data from TCCIP table
## 2.1 established the function
```{r}

tccip_region_ext <- function(i,var,wpath,coord){
  require(data.table)
  dt <- fread(paste0(wpath,"觀測_日資料_臺灣_",var,"/觀測_日資料_臺灣_",var,"_",i,".csv"))
  dt[,c_id:=paste0(LON,LAT)]
  dt_np <-dt[coord,on=.(c_id)]
  dt_np[,c_id:=NULL]
  dt_t <- melt(dt_np,id.vars=c("LON","LAT"),variable.name="date_t")
  dt_t[,date:=as.Date(date_t,format="%Y%m%d")][
    ,c("year","month"):=.(year(date),month(date))]
  dt_t <- dt_t[!is.na(date)]
  if (grepl("溫",var)){
    dt_m <- dt_t[,.(value=mean(value,na.rm=TRUE)),by=.(LON,LAT,year,month)]
  }else{
    dt_m <- dt_t[,.(value=sum(value,na.rm=TRUE)),by=.(LON,LAT,year,month)]
  }
  return(dt_m)
}

```
## 2.2 data processing
```{r}
year <- 1995:2022
core <- makeCluster(10)
temp_mean <-parLapply(core,year,tccip_region_ext,var="平均溫",wpath=w_path,coord=p_coord)
temp_max <-parLapply(core,year,tccip_region_ext,var="最高溫",wpath=w_path,coord=p_coord)
temp_min <-parLapply(core,year,tccip_region_ext,var="最低溫",wpath=w_path,coord=p_coord)
pre <-parLapply(core,year,tccip_region_ext,var="降雨量",wpath=w_path,coord=p_coord)
stopCluster(core)

clm_all <- list(temp_mean,temp_max,temp_min,pre)

```

## 2.3 calculated the weather index

```{r}

clm_all <- lapply(clm_all,rbindlist)
clm_all <- lapply(clm_all,function(clm_dt){
  clm_dt[,wkt:=paste("POLYGON ((",
                   LON-0.005,LAT-0.005,",",
                   LON+0.005,LAT-0.005,",",
                   LON+0.005,LAT+0.005,",",
                   LON-0.005,LAT+0.005,",",
                   LON-0.005,LAT-0.005,"))",
                   sep=" ")]
  return(clm_dt)
})

clm_Y <- lapply(clm_all,function(clm_dt){
  yr <- clm_dt[,.(value=mean(value)),by=.(LON,LAT,year,wkt)]
  yr[,v_mean:=mean(value),by=.(wkt)]
  yr[,v_sd:=sd(value),by=.(wkt)]
  yr[,st_value:=(value-v_mean)/v_sd]
  return(yr)
})
### calculate the annual mean
mean_T <- clm_Y[[1]]

mean_T[,.(mean_t=mean(value)),by=.(wkt)]$mean_t %>% mean()

clm_Y[[4]][,value:=value*12]
y_var <- c("mean_T","max_T"
           ,"min_T","PRE")
lapply(1:3,function(i){
  dt <- clm_Y[[i]]
  var <- y_var[i]
  dir.create(paste0("result/plot/yearly_",var))
  lapply(1995:2022,function(y){
    dt_y <- dt[year==y]
    dt_y <-st_as_sf(dt_y,wkt ="wkt",crs=4326) 
    p <- ggplot()+
          geom_sf(data=dt_y,aes(fill=st_value))+
          labs(fill="STI",title =paste(y,"STI"),sep=" ")+
          scale_fill_gradient2(
            high ="#A62621",
            mid = "gray90",
            low = "#3C56A6",
            midpoint = 0,
            limits=c(-4,4)
          )
    ggsave(paste0("result/plot/yearly_",var,"/",var,"_",y,".jpeg"),
           height = 5,width = 8,dpi=600,plot=p)
  })
})

### create the gif
meanT_imgs_path <- list.files("result/plot/yearly_mean_T/", full.names = TRUE)
meanT_img_list <- lapply(meanT_imgs_path, image_read)  
img_joined <- image_join(meanT_img_list)
img_animated <- image_animate(img_joined, fps = 2)
image_write(image = img_animated,
            path = "result/plot/tccip_year_meanT.gif")
###
y_var <- c("mean_T","max_T"
           ,"min_T","PRE")
clm_Y
lapply(1:3,function(i){
  dt <- clm_Y[[i]]
  var <- y_var[i]
  dir.create(paste0("result/plot/yearly_org_",var))
  lapply(1995:2022,function(y){
    dt_y <- dt[year==y]
    dt_y <-st_as_sf(dt_y,wkt ="wkt",crs=4326) 
    p <- ggplot()+
          geom_sf(data=dt_y,aes(fill=value))+
          labs(fill=var,title =y)+
          scale_fill_gradient2(
            high ="#F97661",
            mid ="gray90",
            low = "#2831DD",
            midpoint = mean(dt$value,na.rm=T),
            limits=c(min(dt$value,na.rm=T),max(dt$value,na.rm=T))
          )
    ggsave(paste0("result/plot/yearly_org_",var,"/org",var,"_",y,".jpeg"),
           height = 5,width = 8,dpi=600,plot=p)
  })
})
PRE_Y <- clm_Y[[4]]
### deal with precipitation
  lapply(1995:2022,function(y){
    dt_y <- PRE_Y[year==y]
    dt_y <-st_as_sf(dt_y,wkt ="wkt",crs=4326) 
    p <- ggplot()+
          geom_sf(data=dt_y,aes(fill=st_value))+
          labs(fill="rPRE",title =y)+
          scale_fill_gradient2(
            high ="#00896C",
            mid = "gray90",
            low = "#CAAD5F",
            midpoint = 0,
            limits=c(-4,4)
          )
    ggsave(paste0("result/plot/yearly_PRE/PRE_",y,".jpeg"),
           height = 5,width = 8,dpi=600,plot=p)
  })
### gif
#### for year_PRE
pre_imgs_path <- list.files("result/plot/yearly_PRE/", full.names = TRUE)
pre_img_list <- lapply(pre_imgs_path, image_read)  
img_joined <- image_join(pre_img_list)
img_animated <- image_animate(img_joined, fps = 2)
image_write(image = img_animated,
            path = "result/plot/tccip_year_PRE.gif")
#### for max_Temp
maxT_imgs_path <- list.files("result/plot/yearly_org_max_T/", full.names = TRUE)
maxT_img_list <- lapply(maxT_imgs_path, image_read) %>% image_join()
img_animated <- image_animate(maxT_img_list, fps = 2)
image_write(image = img_animated,
            path = "result/plot/tccip_yearly_org_max_T.gif")

```
 
## 2.4 save the data
 
```{r}
y_var <- c("mean_T","max_T"
           ,"min_T","PRE")
### save the total weather data table
lapply(1:4,function(i){
  dt <- clm_all[[i]]
  var <- y_var[i]
  fwrite(dt,paste0(pro_path,"rawdata/weather_table/basic_data_",var,".csv"))
  })
```
## 2.5 calculate the long-term temperature variation plot
```{r}
mean_T_y <- mean_T[,.(mean=mean(value),max=max(value),min=min(value)),by=.(year)]
mean_T_y[,STI:=(mean-mean(mean))/sd(mean)]
p1 <- ggplot(data=mean_T_y,aes(x=year,y=mean))+
  geom_point()+
  geom_line()+
  geom_smooth(color="darkred")+
  geom_hline(yintercept = mean(mean_T_y$mean),linetype = "dashed")+
  labs(x="Year",y="Temperature (°C)")+
  scale_x_continuous(breaks = seq(1995,2023,2))+
  theme_bw()

p2 <- ggplot(data=mean_T_y,aes(x=year,y=STI,fill=STI))+
  geom_col()+
  scale_fill_gradient2(
            high ="#A62621",
            mid = "gray90",
            low = "#3C56A6",
            midpoint = 0,
            limits=c(-2,2))+
  labs(x="Year",y="STI")+
  scale_x_continuous(breaks = seq(1995,2023,2))+
  theme_bw()

p1/p2+ plot_annotation(tag_levels = 'A')
ggsave(paste0(pro_path,"result/plot/weather/mean_T_yearly_and_anomaly.jpeg"),width = 8,height = 7)

```
## 2.6 analysis by vegetation zone
```{r}
vz <- st_read(paste0(pro_path,"rawdata/layer/vegetation_zone_vector.gpkg"))
setnames(vz,"DN","VT")
class(mean_T)
fit_temp_l <- mean_T[year==1995] %>% st_as_sf(.,wkt="wkt",crs=4326)

vz_temp <- st_intersection(vz,fit_temp_l)
fit_vt <- vz_temp %>% as.data.table() %>% .[,.(VT,LON,LAT)] %>% unique()
rm(fit_temp_l,vz_temp)
## combine the match table with data
vt_stat <- mean_T[fit_vt,on=.(LON,LAT),allow.cartesian=TRUE] %>% 
  .[,.(temp=mean(value)),by=.(year,VT)]
vt_stat[,.(temp=mean(temp),max=max(temp),min=min(temp)),by=.(VT)] %>%
  write.csv(file = paste0(pro_path,"result/vegetatio_zone_temp_stat.csv"))
vt_stat[,STI:=(temp-mean(temp))/sd(temp),by=.(VT)]
vt_stat[,mean_T:=mean(temp),by=.(VT)]
ggplot(data=vt_stat,aes(x=year,y=temp))+
  geom_line()+
  geom_point()+
  geom_hline(aes(yintercept = mean_T),linetype="dashed")+
  facet_wrap(VT~.,ncol=2,scales = "free_y")+
  geom_smooth(color="darkred")+
  scale_x_continuous(breaks = seq(1995,2023,5))+
  labs(x="Year",y="Temperature (°C)")
ggsave(paste0(pro_path,"result/plot/temp_yearly_by_veg_zone.jpeg"),width=8,height = 9)

ggplot(data=vt_stat,aes(x=year,y=STI,fill=STI))+
  geom_col()+
  facet_wrap(VT~.,ncol=2,scales = "free_y")+
  scale_fill_gradient2(
            high ="#A62621",
            mid = "gray90",
            low = "#3C56A6",
            midpoint = 0,
            limits=c(-2.5,2.5))+
  labs(x="Year",y="Standardized temperature index")+
  scale_x_continuous(breaks = seq(1995,2023,5))
ggsave(paste0(pro_path,"result/plot/STI_yearly_by_veg_zone.jpeg"),width=8,height = 9)
```
## 2.6 precipitation analysis
```{r}
mean(PRE_Y$value)
PRE_Y_stat <- PRE_Y[,.(mean=mean(value)),by=.(year)]
PRE_region <- PRE_Y[,.(value=mean(value)),by=.(wkt)]

PRE_region <-st_as_sf(PRE_region,wkt ="wkt",crs=4326) 
ggplot()+
 geom_sf(data=PRE_region ,aes(fill=value))+
 labs(fill="PRECIP")+
 scale_fill_gradient(
   high ="#145EA4",
   low = "gray90")

ggsave(paste0(pro_path,"result/plot/PRE_mean_region.jpeg"),
      height = 5,width = 8,dpi=600)


```

## 2.7 SPI analysis
```{r}
pre_mon <- clm_all[[4]]
for (i in unique(pre_mon$wkt)){
    pre_mon[wkt==i,value:=ts(value,start = c(1995,1),frequency = 1)]
    pre_mon[wkt==i,spi_3:=as.data.table(spi(value,3)$fitted)][
      wkt==i,spi_12:=as.data.table(spei(value,12)$fitted)]
}
dir.create(paste0(pro_path,"result/plot/weather/SPI/"))
lapply(1996:2022,function(y){
    dt_y <- pre_mon[year==y,.(spi_12=mean(spi_12)),by=.(year,wkt)]
    dt_y <-st_as_sf(dt_y,wkt ="wkt",crs=4326) 
    p <- ggplot()+
          geom_sf(data=dt_y,aes(fill=spi_12))+
          labs(fill="SPI_12",title =y)+
          scale_fill_gradient2(
            high ="#00896C",
            mid = "gray90",
            low = "#CAAD5F",
            midpoint = 0,
            limits=c(-2.5,2.5)
          )
    ggsave(paste0(pro_path,"result/plot/weather/SPI/SPI_",y,".jpeg"),
           height = 5,width = 8,dpi=600,plot=p)
  })
pre_mon_reg <- pre_mon[,.(spi_12=mean(spi_12,na.rm=T),spi_3=mean(spi_3,na.rm=T)),
                       by=.(year)]
ggplot()+
     geom_sf(data=dt_y,aes(fill=spi_12))+
       labs(fill="SPI_12",title =y)+
       scale_fill_gradient2(
                high ="#00896C",
                mid = "gray90",
                low = "#CAAD5F",
                midpoint = 0,
                limits=c(-2.5,2.5))

```


 
# 3. ERA5_data

```{r}
era5_dt <- fread(paste0(pro_path,"rawdata/weather_table/spei_index_data_grid.csv"))
era5_dt[,wkt:=paste("POLYGON ((",
                   x-0.05,y-0.05,",",
                   x+0.05,y-0.05,",",
                   x+0.05,y+0.05,",",
                   x-0.05,y+0.05,",",
                   x-0.05,y-0.05,"))",
                   sep=" ")]
  

era5_monthly_reg <- 
  era5_dt[,.(spei_12=mean(spei_12,na.rm=T),
             spei_24=mean(spei_24,na.rm=T)),by=.(year,month)]
era5_monthly_reg[,date:=as.Date(paste(year,month,"1",sep="-"))]
ggplot(data=era5_monthly_reg,aes(x=date,y=spei_12,fill=spei_12))+
       geom_col()+
       labs(fill="SPEI-12",y="12-month SPEI")+
       scale_fill_gradient2(
            high ="#00896C",
            mid = "gray85",
            low = "#CAAD5F",
         midpoint = 0,
         limits=c(-2.5,2.5)
       )+
  theme_bw()+
  scale_x_date(breaks="2 years",
               limits =c(as.Date("2000-01-01"),as.Date("2024-04-01")),
               date_labels = "%Y")
 ggsave(paste0(pro_path,"result/plot/spei_12_2000_2024_04.jpeg"),
        height = 6,width = 10,dpi=300)

era5_sf_ext <-st_as_sf(era5_dt[year==2021&month==5],wkt ="wkt",crs=4326) 
era5_sf_res<-st_as_sf(era5_dt[year==2024&month==4],wkt ="wkt",crs=4326) 
p1 <- ggplot()+
       geom_sf(data=era5_sf_ext,aes(fill=spei_12),color="gray")+
       geom_sf(data=region,fill=NA)+
       labs(fill="SPEI-12",title ="2021-05")+
       scale_fill_gradient2(
            high ="#00896C",
            mid = "gray95",
            low = "#CAAD5F",
         midpoint = 0,
         limits=c(-2.5,2.5)
       )+
   geom_text(data=era5_sf_ext,aes(x=x,y=y,label=round(spei_12,3)))
p2 <- ggplot()+
       geom_sf(data=era5_sf_res,aes(fill=spei_12),color="gray")+
       geom_sf(data=region,fill=NA)+
       labs(fill="SPEI-12",title ="2024-04")+
       scale_fill_gradient2(
            high ="#00896C",
            mid = "gray95",
            low = "#CAAD5F",
         midpoint = 0,
         limits=c(-2.5,2.5)
       )+
   geom_text(data=era5_sf_res,aes(x=x,y=y,label=round(spei_12,3)))
p1/p2+ plot_layout(guides = "collect")+ plot_annotation(tag_levels = 'A')

ggsave(paste0(pro_path,"result/plot/spei_12_region_display.jpeg"),
        height = 10,width = 8,dpi=600)
p1+p2+ plot_layout(guides = "collect")+ plot_annotation(tag_levels = 'A')
ggsave(paste0(pro_path,"result/plot/spei_12_region_display_2.jpeg"),
        height = 10,width = 8,dpi=600)

```

